### 2025-05-12 writeup/paper.tex
added paper format

% This must be in the first 5 lines to tell arXiv to use pdfLaTeX, which is strongly recommended.
\pdfoutput=1
% In particular, the hyperref package requires pdfLaTeX in order to break URLs across lines.

\documentclass[11pt]{article}

% Change "review" to "final" to generate the final (sometimes called camera-ready) version.
% Change to "preprint" to generate a non-anonymous version with page numbers.
\usepackage[review]{acl}

% Standard package includes
\usepackage{times}
\usepackage{latexsym}

% For proper rendering and hyphenation of words containing Latin characters (including in bib files)
\usepackage[T1]{fontenc}
% For Vietnamese characters
% \usepackage[T5]{fontenc}
% See https://www.latex-project.org/help/documentation/encguide.pdf for other character sets

% This assumes your files are encoded as UTF8
\usepackage[utf8]{inputenc}

% This is not strictly necessary, and may be commented out,
% but it will improve the layout of the manuscript,
% and will typically save some space.
\usepackage{microtype}

% This is also not strictly necessary, and may be commented out.
% However, it will improve the aesthetics of text in
% the typewriter font.
\usepackage{inconsolata}

%Including images in your LaTeX document requires adding
%additional package(s)
\usepackage{graphicx}

% If the title and author information does not fit in the area allocated, uncomment the following
%
%\setlength\titlebox{<dim>}
%
% and set <dim> to something 5cm or larger.

\title{Instructions for *ACL Proceedings}

% Author information can be set in various styles:
% For several authors from the same institution:
% \author{Author 1 \and ... \and Author n \\
%         Address line \\ ... \\ Address line}
% if the names do not fit well on one line use
%         Author 1 \\ {\bf Author 2} \\ ... \\ {\bf Author n} \\
% For authors from different institutions:
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \And  ... \And
%         Author n \\ Address line \\ ... \\ Address line}
% To start a separate ``row'' of authors use \AND, as in
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \AND
%         Author 2 \\ Address line \\ ... \\ Address line \And
%         Author 3 \\ Address line \\ ... \\ Address line}

\author{First Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  \texttt{email@domain} \\\And
  Second Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  \texttt{email@domain} \\}

%\author{
%  \textbf{First Author\textsuperscript{1}},
%  \textbf{Second Author\textsuperscript{1,2}},
%  \textbf{Third T. Author\textsuperscript{1}},
%  \textbf{Fourth Author\textsuperscript{1}},
%\\
%  \textbf{Fifth Author\textsuperscript{1,2}},
%  \textbf{Sixth Author\textsuperscript{1}},
%  \textbf{Seventh Author\textsuperscript{1}},
%  \textbf{Eighth Author \textsuperscript{1,2,3,4}},
%\\
%  \textbf{Ninth Author\textsuperscript{1}},
%  \textbf{Tenth Author\textsuperscript{1}},
%  \textbf{Eleventh E. Author\textsuperscript{1,2,3,4,5}},
%  \textbf{Twelfth Author\textsuperscript{1}},
%\\
%  \textbf{Thirteenth Author\textsuperscript{3}},
%  \textbf{Fourteenth F. Author\textsuperscript{2,4}},
%  \textbf{Fifteenth Author\textsuperscript{1}},
%  \textbf{Sixteenth Author\textsuperscript{1}},
%\\
%  \textbf{Seventeenth S. Author\textsuperscript{4,5}},
%  \textbf{Eighteenth Author\textsuperscript{3,4}},
%  \textbf{Nineteenth N. Author\textsuperscript{2,5}},
%  \textbf{Twentieth Author\textsuperscript{1}}
%\\
%\\
%  \textsuperscript{1}Affiliation 1,
%  \textsuperscript{2}Affiliation 2,
%  \textsuperscript{3}Affiliation 3,
%  \textsuperscript{4}Affiliation 4,
%  \textsuperscript{5}Affiliation 5
%\\
%  \small{
%    \textbf{Correspondence:} \href{mailto:email@domain}{email@domain}
%  }
%}

\begin{document}
\maketitle
\begin{abstract}
This document is a supplement to the general instructions for *ACL authors. It contains instructions for using the \LaTeX{} style files for ACL conferences.
The document itself conforms to its own specifications, and is therefore an example of what your manuscript should look like.
These instructions should be used both for papers submitted for review and for final versions of accepted papers.
\end{abstract}

\section{Introduction}

These instructions are for authors submitting papers to *ACL conferences using \LaTeX. They are not self-contained. All authors must follow the general instructions for *ACL proceedings,\footnote{\url{http://acl-org.github.io/ACLPUB/formatting.html}} and this document contains additional instructions for the \LaTeX{} style files.

The templates include the \LaTeX{} source of this document (\texttt{acl\_latex.tex}),
the \LaTeX{} style file used to format it (\texttt{acl.sty}),
an ACL bibliography style (\texttt{acl\_natbib.bst}),
an example bibliography (\texttt{custom.bib}),
and the bibliography for the ACL Anthology (\texttt{anthology.bib}).

\section{Engines}

To produce a PDF file, pdf\LaTeX{} is strongly recommended (over original \LaTeX{} plus dvips+ps2pdf or dvipdf).
The style file \texttt{acl.sty} can also be used with
lua\LaTeX{} and
Xe\LaTeX{}, which are especially suitable for text in non-Latin scripts.
The file \texttt{acl\_lualatex.tex} in this repository provides
an example of how to use \texttt{acl.sty} with either
lua\LaTeX{} or
Xe\LaTeX{}.

\section{Preamble}

The first line of the file must be
\begin{quote}
\begin{verbatim}
\documentclass[11pt]{article}
\end{verbatim}
\end{quote}

To load the style file in the review version:
\begin{quote}
\begin{verbatim}
\usepackage[review]{acl}
\end{verbatim}
\end{quote}
For the final version, omit the \verb|review| option:
\begin{quote}
\begin{verbatim}
\usepackage{acl}
\end{verbatim}
\end{quote}

To use Times Roman, put the following in the preamble:
\begin{quote}
\begin{verbatim}
\usepackage{times}
\end{verbatim}
\end{quote}
(Alternatives like txfonts or newtx are also acceptable.)

Please see the \LaTeX{} source of this document for comments on other packages that may be useful.

Set the title and author using \verb|\title| and \verb|\author|. Within the author list, format multiple authors using \verb|\and| and \verb|\And| and \verb|\AND|; please see the \LaTeX{} source for examples.

By default, the box containing the title and author names is set to the minimum of 5 cm. If you need more space, include the following in the preamble:
\begin{quote}
\begin{verbatim}
\setlength\titlebox{<dim>}
\end{verbatim}
\end{quote}
where \verb|<dim>| is replaced with a length. Do not set this length smaller than 5 cm.

\section{Document Body}

\subsection{Footnotes}

Footnotes are inserted with the \verb|\footnote| command.\footnote{This is a footnote.}

\subsection{Tables and figures}

See Table~\ref{tab:accents} for an example of a table and its caption.
\textbf{Do not override the default caption sizes.}

\begin{table}
  \centering
  \begin{tabular}{lc}
    \hline
    \textbf{Command} & \textbf{Output} \\
    \hline
    \verb|{\"a}|     & {\"a}           \\
    \verb|{\^e}|     & {\^e}           \\
    \verb|{\`i}|     & {\`i}           \\
    \verb|{\.I}|     & {\.I}           \\
    \verb|{\o}|      & {\o}            \\
    \verb|{\'u}|     & {\'u}           \\
    \verb|{\aa}|     & {\aa}           \\\hline
  \end{tabular}
  \begin{tabular}{lc}
    \hline
    \textbf{Command} & \textbf{Output} \\
    \hline
    \verb|{\c c}|    & {\c c}          \\
    \verb|{\u g}|    & {\u g}          \\
    \verb|{\l}|      & {\l}            \\
    \verb|{\~n}|     & {\~n}           \\
    \verb|{\H o}|    & {\H o}          \\
    \verb|{\v r}|    & {\v r}          \\
    \verb|{\ss}|     & {\ss}           \\
    \hline
  \end{tabular}
  \caption{Example commands for accented characters, to be used in, \emph{e.g.}, Bib\TeX{} entries.}
  \label{tab:accents}
\end{table}

As much as possible, fonts in figures should conform
to the document fonts. See Figure~\ref{fig:experiments} for an example of a figure and its caption.

Using the \verb|graphicx| package graphics files can be included within figure
environment at an appropriate point within the text.
The \verb|graphicx| package supports various optional arguments to control the
appearance of the figure.
You must include it explicitly in the \LaTeX{} preamble (after the
\verb|\documentclass| declaration and before \verb|\begin{document}|) using
\verb|\usepackage{graphicx}|.

\begin{figure}[t]
  \includegraphics[width=\columnwidth]{example-image-golden}
  \caption{A figure with a caption that runs for more than one line.
    Example image is usually available through the \texttt{mwe} package
    without even mentioning it in the preamble.}
  \label{fig:experiments}
\end{figure}

\begin{figure*}[t]
  \includegraphics[width=0.48\linewidth]{example-image-a} \hfill
  \includegraphics[width=0.48\linewidth]{example-image-b}
  \caption {A minimal working example to demonstrate how to place
    two images side-by-side.}
\end{figure*}

\subsection{Hyperlinks}

Users of older versions of \LaTeX{} may encounter the following error during compilation:
\begin{quote}
\verb|\pdfendlink| ended up in different nesting level than \verb|\pdfstartlink|.
\end{quote}
This happens when pdf\LaTeX{} is used and a citation splits across a page boundary. The best way to fix this is to upgrade \LaTeX{} to 2018-12-01 or later.

\subsection{Citations}

\begin{table*}
  \centering
  \begin{tabular}{lll}
    \hline
    \textbf{Output}           & \textbf{natbib command} & \textbf{ACL only command} \\
    \hline
    \citep{Gusfield:97}       & \verb|\citep|           &                           \\
    \citealp{Gusfield:97}     & \verb|\citealp|         &                           \\
    \citet{Gusfield:97}       & \verb|\citet|           &                           \\
    \citeyearpar{Gusfield:97} & \verb|\citeyearpar|     &                           \\
    \citeposs{Gusfield:97}    &                         & \verb|\citeposs|          \\
    \hline
  \end{tabular}
  \caption{\label{citation-guide}
    Citation commands supported by the style file.
    The style is based on the natbib package and supports all natbib citation commands.
    It also supports commands defined in previous ACL style files for compatibility.
  }
\end{table*}

Table~\ref{citation-guide} shows the syntax supported by the style files.
We encourage you to use the natbib styles.
You can use the command \verb|\citet| (cite in text) to get ``author (year)'' citations, like this citation to a paper by \citet{Gusfield:97}.
You can use the command \verb|\citep| (cite in parentheses) to get ``(author, year)'' citations \citep{Gusfield:97}.
You can use the command \verb|\citealp| (alternative cite without parentheses) to get ``author, year'' citations, which is useful for using citations within parentheses (e.g. \citealp{Gusfield:97}).

A possessive citation can be made with the command \verb|\citeposs|.
This is not a standard natbib command, so it is generally not compatible
with other style files.

\subsection{References}

\nocite{Ando2005,andrew2007scalable,rasooli-tetrault-2015}

The \LaTeX{} and Bib\TeX{} style files provided roughly follow the American Psychological Association format.
If your own bib file is named \texttt{custom.bib}, then placing the following before any appendices in your \LaTeX{} file will generate the references section for you:
\begin{quote}
\begin{verbatim}
\bibliography{custom}
\end{verbatim}
\end{quote}

You can obtain the complete ACL Anthology as a Bib\TeX{} file from \url{https://aclweb.org/anthology/anthology.bib.gz}.
To include both the Anthology and your own .bib file, use the following instead of the above.
\begin{quote}
\begin{verbatim}
\bibliography{anthology,custom}
\end{verbatim}
\end{quote}

Please see Section~\ref{sec:bibtex} for information on preparing Bib\TeX{} files.

\subsection{Equations}

An example equation is shown below:
\begin{equation}
  \label{eq:example}
  A = \pi r^2
\end{equation}

Labels for equation numbers, sections, subsections, figures and tables
are all defined with the \verb|\label{label}| command and cross references
to them are made with the \verb|\ref{label}| command.

This an example cross-reference to Equation~\ref{eq:example}.

\subsection{Appendices}

Use \verb|\appendix| before any appendix section to switch the section numbering over to letters. See Appendix~\ref{sec:appendix} for an example.

\section{Bib\TeX{} Files}
\label{sec:bibtex}

Unicode cannot be used in Bib\TeX{} entries, and some ways of typing special characters can disrupt Bib\TeX's alphabetization. The recommended way of typing special characters is shown in Table~\ref{tab:accents}.

Please ensure that Bib\TeX{} records contain DOIs or URLs when possible, and for all the ACL materials that you reference.
Use the \verb|doi| field for DOIs and the \verb|url| field for URLs.
If a Bib\TeX{} entry has a URL or DOI field, the paper title in the references section will appear as a hyperlink to the paper, using the hyperref \LaTeX{} package.

\section*{Limitations}

Since December 2023, a "Limitations" section has been required for all papers submitted to ACL Rolling Review (ARR). This section should be placed at the end of the paper, before the references. The "Limitations" section (along with, optionally, a section for ethical considerations) may be up to one page and will not count toward the final page limit. Note that these files may be used by venues that do not rely on ARR so it is recommended to verify the requirement of a "Limitations" section and other criteria with the venue in question.

\section*{Acknowledgments}

This document has been adapted
by Steven Bethard, Ryan Cotterell and Rui Yan
from the instructions for earlier ACL and NAACL proceedings, including those for
ACL 2019 by Douwe Kiela and Ivan Vuli\'{c},
NAACL 2019 by Stephanie Lukin and Alla Roskovskaya,
ACL 2018 by Shay Cohen, Kevin Gimpel, and Wei Lu,
NAACL 2018 by Margaret Mitchell and Stephanie Lukin,
Bib\TeX{} suggestions for (NA)ACL 2017/2018 from Jason Eisner,
ACL 2017 by Dan Gildea and Min-Yen Kan,
NAACL 2017 by Margaret Mitchell,
ACL 2012 by Maggie Li and Michael White,
ACL 2010 by Jing-Shin Chang and Philipp Koehn,
ACL 2008 by Johanna D. Moore, Simone Teufel, James Allan, and Sadaoki Furui,
ACL 2005 by Hwee Tou Ng and Kemal Oflazer,
ACL 2002 by Eugene Charniak and Dekang Lin,
and earlier ACL and EACL formats written by several people, including
John Chen, Henry S. Thompson and Donald Walker.
Additional elements were taken from the formatting instructions of the \emph{International Joint Conference on Artificial Intelligence} and the \emph{Conference on Computer Vision and Pattern Recognition}.

% Bibliography entries for the entire Anthology, followed by custom entries
%\bibliography{anthology,custom}
% Custom bibliography entries only
\bibliography{custom}

\appendix

\section{Example Appendix}
\label{sec:appendix}

This is an appendix.

\end{document}

### 2025-05-12 writeup/paper.tex
title and authors

@@
 % Change to "preprint" to generate a non-anonymous version with page numbers.
-\usepackage[review]{acl}
+% \usepackage[review]{acl}
+\usepackage[preprint]{acl}
 
@@
 
-\title{Instructions for *ACL Proceedings}
+\title{Are Modern Byte Pair Tokenizers Zipfian?}
 
@@
 
-\author{First Author \\
-  Affiliation / Address line 1 \\
-  Affiliation / Address line 2 \\
-  Affiliation / Address line 3 \\
-  \texttt{email@domain} \\\And
-  Second Author \\
-  Affiliation / Address line 1 \\
-  Affiliation / Address line 2 \\
-  Affiliation / Address line 3 \\
-  \texttt{email@domain} \\}
+\author{Jack Hanke \\
+  Northwestern University \\
+  \And
+  Daniel Plotkin \\
+  Northwestern University \\
+  \And
+  Nicole Birova \\
+  Northwestern University}
 

### 2025-05-14 writeup/paper.tex
whole draft lol

@@
 
+% Jack!
+% \usepackage{amsmath}
+
 % If the title and author information does not fit in the area allocated, uncomment the following
@@
 
-\title{Are Modern Byte Pair Tokenizers Zipfian?}
+\title{Modern Byte Pair Tokenizers are Zipfian}
 
@@
 \begin{abstract}
-This document is a supplement to the general instructions for *ACL authors. It contains instructions for using the \LaTeX{} style files for ACL conferences.
-The document itself conforms to its own specifications, and is therefore an example of what your manuscript should look like.
-These instructions should be used both for papers submitted for review and for final versions of accepted papers.
+A majority of large language models ingest word fragments called \textit{tokens} produced by a data compression algorithm known as byte pair encoding.
+This algorithm groups high-frequency letter pairings in natural language into individual units.
+A natural question is whether the tokens produced by this pairing process deviate significantly from the source language's frequency distribution.
+Zipf famously showed that many natural language's frequency distribution follows a power law, commonly known as Zipf's Law.
+We examine two modern tokenizer's adherence to Zipf's law at the token level.
+We demonstrate that these tokenizers are Zipfian on two corpuses, and speculate as to why this is.
 \end{abstract}
@@
 
-These instructions are for authors submitting papers to *ACL conferences using \LaTeX.
-They are not self-contained.
-All authors must follow the general instructions for *ACL proceedings,\footnote{\url{http://acl-org.github.io/ACLPUB/formatting.html}} and this document contains additional instructions for the \LaTeX{} style files.
+George Zipf in his hallmark \textit{The Psycho-Biology of Language} \cite{Zip35} introduced the remarkable trend that the word frequencies in many human languages exhibit the same power law distribution.
+Specifically, Zipf said that the frequency  that a word appears in some large corpus is proportional to the word's frequency rank.
+Mathematically, Zipf's laws states
 
-The templates include the \LaTeX{} source of this document (\texttt{acl\_latex.tex}),
-the \LaTeX{} style file used to format it (\texttt{acl.sty}),
-an ACL bibliography style (\texttt{acl\_natbib.bst}),
-an example bibliography (\texttt{custom.bib}),
-and the bibliography for the ACL Anthology (\texttt{anthology.bib}).
+\begin{equation}
+    \mbox{word frequency} \propto \frac{1}{\mbox{word rank}}.
+\end{equation}
 
-\section{Engines}
+A more descriptive model of word frequency that is more commonly referenced in linguistics literature is the Zipf-Mandelbrot distribution
 
-To produce a PDF file, pdf\LaTeX{} is strongly recommended (over original \LaTeX{} plus dvips+ps2pdf or dvipdf).
-The style file \texttt{acl.sty} can also be used with
-lua\LaTeX{} and
-Xe\LaTeX{}, which are especially suitable for text in non-Latin scripts.
-The file \texttt{acl\_lualatex.tex} in this repository provides
-an example of how to use \texttt{acl.sty} with either
-lua\LaTeX{} or
-Xe\LaTeX{}.
+\begin{equation}
+    \mbox{word frequency} \propto \frac{1}{(\mbox{word rank} + b)^a}.
+    \label{eq:zipfmandelbrot}
+\end{equation}
 
-\section{Preamble}
+where $a,b$ are fitted parameters. We say a distribution that follows the trend in Equation \ref{eq:zipfmandelbrot} with $a \approx 1$ is \textit{Zipf distributed}, or simply \textit{Zipfian}.
 
-The first line of the file must be
-\begin{quote}
-\begin{verbatim}
-\documentclass[11pt]{article}
-\end{verbatim}
-\end{quote}
+The accuracy of the trend in Equation \ref{eq:zipfmandelbrot} has been examined in $10^8$ English words in \cite{cancho2000}, over 50 languages in \cite{yu2018zipfslaw50languages}, and written-versus-spoken corpuses in \cite{lin2015scalinglawshumanspeech}.
+Each of these studies demonstrate that Zipf's law is generally exhibitted for common and somewhat-uncommon words, but rare words (words with high rank) appear less frequently than predicted.
+This deviation is shown to be statistically significant, and appears as two trendlines in the log-log plot.
+In English corpuses, this transition is found around the $10^5$\textit{-th} ranked word.\footnote{Some of these studies also find that extremely common words appear slighly more commonly than the Zipfian prediction.} The authors further explore the linguistic relavence and universality of these multiple regimes of words.
 
-To load the style file in the review version:
-\begin{quote}
-\begin{verbatim}
-\usepackage[review]{acl}
-\end{verbatim}
-\end{quote}
-For the final version, omit the \verb|review| option:
-\begin{quote}
-\begin{verbatim}
-\usepackage{acl}
-\end{verbatim}
-\end{quote}
+Nearly a century after Zipf's discovery, large language models (LLMs) generate text comparable to human communication.
+However, unlike humans, LLMs digest text using a fixed vocabulary of word fragments called \textit{tokens}.
+The mapping between natural language and tokens is most commonly computed using the \textit{byte pair encoding algorithm} \cite{bpegage} over some large training corpus.
+Byte pair encoding identifies the most frequent pair of letters, and creates a new token in the "token vocabulary" to replace that pair.
+Iterating this procedure until some fixed vocabulary size is reached creates the vocabulary of tokens.
+Finally, any sequence that does not map to one of the derived tokens is labelled with the \texttt{<unk>} token, standing for unknown.
 
-To use Times Roman, put the following in the preamble:
-\begin{quote}
-\begin{verbatim}
-\usepackage{times}
-\end{verbatim}
-\end{quote}
-(Alternatives like txfonts or newtx are also acceptable.)
+When considering these ideas in tandem, a natural question arises: given a corpus that appears Zipfian, how does the tokenization process affect this trend?
 
-Please see the \LaTeX{} source of this document for comments on other packages that may be useful.
+\section{Methods}
 
-Set the title and author using \verb|\title| and \verb|\author|.
-Within the author list, format multiple authors using \verb|\and| and \verb|\And| and \verb|\AND|; please see the \LaTeX{} source for examples.
+To explore this question, we choose two modern tokenizers to analyze.
+We choose the tokenizer for the RoBERTa language model \cite{liu2019robertarobustlyoptimizedbert}, as the training data is publicly known.
+This allows us to conduct frequency analyses on corpuses that are on and off-distribution for the tokenizer.
+We also choose to compare our results with the tokenizer for the GPT-4 language model \cite{openai2024gpt4technicalreport} as a representative of an industry-grade tokenizer.
 
-By default, the box containing the title and author names is set to the minimum of 5 cm. If you need more space, include the following in the preamble:
-\begin{quote}
-\begin{verbatim}
-\setlength\titlebox{<dim>}
-\end{verbatim}
-\end{quote}
-where \verb|<dim>| is replaced with a length. Do not set this length smaller than 5 cm.
+We choose two corpuses based on RoBERTa's training data.
+We choose the 4.4GB bookscorpus dataset \cite{zhu2015aligningbooksmoviesstorylike}, as RoBERTa's tokenizer was trained (in-part) on this corpus.
+For our off-distribution corpus, we choose the 5.6GBtying MiniPile dataset \cite{kaddour2023minipilechallengedataefficientlanguage}, which was released after the RoBERTa model was released.
+Note that it is likely that the GPT-4 tokenizer was trained (in-part) on both of these corpuses.
 
-\section{Document Body}
+We compute the word and token frequency for both tokenizers on each corpus, where we define a word as anything separated by a space.
+We then compute the data's Kolmogorov-Smirnov goodness-of-fit statistic for the Zipf distribution, and compare to that for an exponential and log-normal distribution.
+We do not remove any tokens from consideration unless otherwise stated.
 
-\subsection{Footnotes}
+\section{Results}
 
-Footnotes are inserted with the \verb|\footnote| command.\footnote{This is a footnote.}
+We find that token vocabularies are remarkably Zipfian, excluding extremely rare tokens.
+This holds true for both tokenizers and both corpuses studied, including out-of-distribution text.
+In the log-log plots in Figure \ref{fig:experiments}, we show sharp declines from the Zipfian trend only for the final few tokens in the RoBERTa tokenizer for both corpuses and the GPT-4 tokenizer for the MiniPile corpus.
+However, the GPT-4 tokenizer trend on the bookscorpus seems to deviate more dramatically, similar to the multiple regime studies.
 
-\subsection{Tables and figures}
+\begin{figure}[t]
+  \includegraphics[width=\columnwidth]{../visualizations/bookscorpusfreq.png}
+  \includegraphics[width=\columnwidth]{../visualizations/minipilecorpusfreq.png}
+  \caption{Log-log plots of word \& token rank vs word \& token frequency}
+  \label{fig:experiments}
+\end{figure}
 
-See Table~\ref{tab:accents} for an example of a table and its caption.
-\textbf{Do not override the default caption sizes.}
+The lowest rank words and tokens for each tokenizer and each corpus, including control tokens and punctuation, are summarized in Table \ref{tbl:token ranks}.
 
@@
     \centering
-  \begin{tabular}{lc}
+    \begin{tabular}{|c|c|c|c|c|c|c|}
         \hline
-    \textbf{Command} & \textbf{Output} \\
+        \textbf{Rank} & \multicolumn{2}{c|}{\textbf{Words}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4}}  \\
         \hline
-    \verb|{\"a}|     & {\"a}           \\
-    \verb|{\^e}|     & {\^e}           \\
-    \verb|{\`i}|     & {\`i}           \\
-    \verb|{\.I}|     & {\.I}           \\
-    \verb|{\o}|      & {\o}            \\
-    \verb|{\'u}|     & {\'u}           \\
-    \verb|{\aa}|     & {\aa}           \\\hline
-  \end{tabular}
-  \begin{tabular}{lc}
+        1 & . & the & <s> & ␣ & `` & , \\
         \hline
-    \textbf{Command} & \textbf{Output} \\
+        2 & , & of & \textbackslash n & \textbackslash n & i & ␣the \\
         \hline
-    \verb|{\c c}|    & {\c c}          \\
-    \verb|{\u g}|    & {\u g}          \\
-    \verb|{\l}|      & {\l}            \\
-    \verb|{\~n}|     & {\~n}           \\
-    \verb|{\H o}|    & {\H o}          \\
-    \verb|{\v r}|    & {\v r}          \\
-    \verb|{\ss}|     & {\ss}           \\
+        3 & the & and & <\textbackslash s> & . & he & . \\
         \hline
     \end{tabular}
-  \caption{Example commands for accented characters, to be used in, \emph{e.g.}, Bib\TeX{} entries.}
-  \label{tab:accents}
+    \caption{Summary of most common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn MiniPile, including control tokens and punctuation}
+    \label{tbl:token ranks}
 \end{table}
 
-As much as possible, fonts in figures should conform
-to the document fonts. See Figure~\ref{fig:experiments} for an example of a figure and its caption.
-
-Using the \verb|graphicx| package graphics files can be included within figure
-environment at an appropriate point within the text.
-The \verb|graphicx| package supports various optional arguments to control the
-appearance of the figure.
-You must include it explicitly in the \LaTeX{} preamble (after the
-\verb|\documentclass| declaration and before \verb|\begin{document}|) using
-\verb|\usepackage{graphicx}|.
-
-\begin{figure}[t]
-  \includegraphics[width=\columnwidth]{example-image-golden}
-  \caption{A figure with a caption that runs for more than one line.
-    Example image is usually available through the \texttt{mwe} package
-    without even mentioning it in the preamble.}
-  \label{fig:experiments}
-\end{figure}
-
-\begin{figure*}[t]
-  \includegraphics[width=0.48\linewidth]{example-image-a} \hfill
-  \includegraphics[width=0.48\linewidth]{example-image-b}
-  \caption {A minimal working example to demonstrate how to place
-    two images side-by-side.}
-\end{figure*}
-
-\subsection{Hyperlinks}
-
-Users of older versions of \LaTeX{} may encounter the following error during compilation:
-\begin{quote}
-\verb|\pdfendlink| ended up in different nesting level than \verb|\pdfstartlink|.
-\end{quote}
-This happens when pdf\LaTeX{} is used and a citation splits across a page boundary. The best way to fix this is to upgrade \LaTeX{} to 2018-12-01 or later.
-
-\subsection{Citations}
-
-\begin{table*}
+\begin{table}
     \centering
-  \begin{tabular}{lll}
+    \begin{tabular}{|c|c|c|c|c|c|c|}
         \hline
-    \textbf{Output}           & \textbf{natbib command} & \textbf{ACL only command} \\
+        \textbf{Rank} & \multicolumn{2}{c|}{\textbf{Words}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4}}  \\
         \hline
-    \citep{Gusfield:97}       & \verb|\citep|           &                           \\
-    \citealp{Gusfield:97}     & \verb|\citealp|         &                           \\
-    \citet{Gusfield:97}       & \verb|\citet|           &                           \\
-    \citeyearpar{Gusfield:97} & \verb|\citeyearpar|     &                           \\
-    \citeposs{Gusfield:97}    &                         & \verb|\citeposs|          \\
+        1 & the & the & the & the & i & the \\
+        \hline
+        2 & to & of & to & of & he & of \\
+        \hline
+        3 & i & and & and & and & she & and \\
         \hline
     \end{tabular}
-  \caption{\label{citation-guide}
-    Citation commands supported by the style file.
-    The style is based on the natbib package and supports all natbib citation commands.
-    It also supports commands defined in previous ACL style files for compatibility.
-  }
-\end{table*}
+    \caption{Summary of most common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn MiniPile, \textit{not} including control tokens and punctuation}
+    \label{tbl:token ranks no control}
+\end{table}
 
-Table~\ref{citation-guide} shows the syntax supported by the style files.
-We encourage you to use the natbib styles.
-You can use the command \verb|\citet| (cite in text) to get ``author (year)'' citations, like this citation to a paper by \citet{Gusfield:97}.
-You can use the command \verb|\citep| (cite in parentheses) to get ``(author, year)'' citations \citep{Gusfield:97}.
-You can use the command \verb|\citealp| (alternative cite without parentheses) to get ``author, year'' citations, which is useful for using citations within parentheses (e.g. \citealp{Gusfield:97}).
+The p-values for the Kolmogorov-Smirnov Tests for each considered distribution are summarized in Table \ref{tbl:pvals}.
 
-A possessive citation can be made with the command \verb|\citeposs|.
-This is not a standard natbib command, so it is generally not compatible
-with other style files.
+\begin{table}
+    \centering
+    \begin{tabular}{|c|c|c|}
+        \hline
+        \textbf{Distribution} & \textbf{bookscorpus} & \textbf{MiniPile} \\
+        \hline
+        Zipf &  &  \\
+        \hline
+        Exponential &  &  \\
+        \hline
+        log-normal &  &  \\
+        \hline
+    \end{tabular}
+    \caption{RoBERTa (GPT-4) tokenizer's Kolmogorov-Smirnov test p-values}
+    \label{tbl:pvals}
+\end{table}
 
-\subsection{References}
+\section{Conclusions}
 
-\nocite{Ando2005,andrew2007scalable,rasooli-tetrault-2015}
+Despite
 
-The \LaTeX{} and Bib\TeX{} style files provided roughly follow the American Psychological Association format.
-If your own bib file is named \texttt{custom.bib}, then placing the following before any appendices in your \LaTeX{} file will generate the references section for you:
-\begin{quote}
-\begin{verbatim}
-\bibliography{custom}
-\end{verbatim}
-\end{quote}
+\section{Future Work}
 
-You can obtain the complete ACL Anthology as a Bib\TeX{} file from \url{https://aclweb.org/anthology/anthology.bib.gz}.
-To include both the Anthology and your own .bib file, use the following instead of the above.
-\begin{quote}
-\begin{verbatim}
-\bibliography{anthology,custom}
-\end{verbatim}
-\end{quote}
+TODO
 
-Please see Section~\ref{sec:bibtex} for information on preparing Bib\TeX{} files.
-
-\subsection{Equations}
-
-An example equation is shown below:
-\begin{equation}
-  \label{eq:example}
-  A = \pi r^2
-\end{equation}
-
-Labels for equation numbers, sections, subsections, figures and tables
-are all defined with the \verb|\label{label}| command and cross references
-to them are made with the \verb|\ref{label}| command.
-
-This an example cross-reference to Equation~\ref{eq:example}.
-
-\subsection{Appendices}
-
-Use \verb|\appendix| before any appendix section to switch the section numbering over to letters. See Appendix~\ref{sec:appendix} for an example.
-
-\section{Bib\TeX{} Files}
-\label{sec:bibtex}
-
-Unicode cannot be used in Bib\TeX{} entries, and some ways of typing special characters can disrupt Bib\TeX's alphabetization.
-The recommended way of typing special characters is shown in Table~\ref{tab:accents}.
-
-Please ensure that Bib\TeX{} records contain DOIs or URLs when possible, and for all the ACL materials that you reference.
-Use the \verb|doi| field for DOIs and the \verb|url| field for URLs.
-If a Bib\TeX{} entry has a URL or DOI field, the paper title in the references section will appear as a hyperlink to the paper, using the hyperref \LaTeX{} package.
-
-\section*{Limitations}
-
-Since December 2023, a "Limitations" section has been required for all papers submitted to ACL Rolling Review (ARR).
-This section should be placed at the end of the paper, before the references.
-The "Limitations" section (along with, optionally, a section for ethical considerations) may be up to one page and will not count toward the final page limit.
-Note that these files may be used by venues that do not rely on ARR so it is recommended to verify the requirement of a "Limitations" section and other criteria with the venue in question.
-
-\section*{Acknowledgments}
-
-This document has been adapted
-by Steven Bethard, Ryan Cotterell and Rui Yan
-from the instructions for earlier ACL and NAACL proceedings, including those for
-ACL 2019 by Douwe Kiela and Ivan Vuli\'{c},
-NAACL 2019 by Stephanie Lukin and Alla Roskovskaya,
-ACL 2018 by Shay Cohen, Kevin Gimpel, and Wei Lu,
-NAACL 2018 by Margaret Mitchell and Stephanie Lukin,
-Bib\TeX{} suggestions for (NA)ACL 2017/2018 from Jason Eisner,
-ACL 2017 by Dan Gildea and Min-Yen Kan,
-NAACL 2017 by Margaret Mitchell,
-ACL 2012 by Maggie Li and Michael White,
-ACL 2010 by Jing-Shin Chang and Philipp Koehn,
-ACL 2008 by Johanna D. Moore, Simone Teufel, James Allan, and Sadaoki Furui,
-ACL 2005 by Hwee Tou Ng and Kemal Oflazer,
-ACL 2002 by Eugene Charniak and Dekang Lin,
-and earlier ACL and EACL formats written by several people, including
-John Chen, Henry S. Thompson and Donald Walker.
-Additional elements were taken from the formatting instructions of the \emph{International Joint Conference on Artificial Intelligence} and the \emph{Conference on Computer Vision and Pattern Recognition}.
+How do token vocabularies differ from english vocabularies? Fixed vocabulary size, catchall token, functional tokens
 
@@
 
-\appendix
+% \appendix
 
-\section{Example Appendix}
-\label{sec:appendix}
+% \section{Example Appendix}
+% \label{sec:appendix}
 
-This is an appendix.
+% This is an appendix.
 

### 2025-05-14 writeup/paper.tex
kstests

@@
 
-The p-values for the Kolmogorov-Smirnov Tests for each considered distribution are summarized in Table \ref{tbl:pvals}.
+The Kolmogorov-Smirnov test statistics for each considered distribution are summarized in Table \ref{tbl:kstest}.
 
@@
     \centering
-    \begin{tabular}{|c|c|c|}
+    \begin{tabular}{|c|c|c|c|c|c|c|}
         \hline
-        \textbf{Distribution} & \textbf{bookscorpus} & \textbf{MiniPile} \\
+        \textbf{Distribution} & \multicolumn{2}{c|}{\textbf{Words}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4}}  \\
         \hline
-        Zipf &  &  \\
+        Zipf &  &  &  &  &  &  \\
         \hline
-        Exponential &  &  \\
+        Exponential & 0.874 & 0.808 & 0.580 & 0.482 & 0.718 & 0.563 \\
         \hline
-        log-normal &  &  \\
+        log-normal & 0.379 & 0.451 & 0.032 & 0.050 & 0.397 & 0.283 \\
         \hline
     \end{tabular}
-    \caption{RoBERTa (GPT-4) tokenizer's Kolmogorov-Smirnov test p-values}
-    \label{tbl:pvals}
+    \caption{Kolmogorov-Smirnov test statistics for distribution fit, where the left subcolumn is bookscorpus and the right subcolumn MiniPile}
+    \label{tbl:kstest}
 \end{table}

### 2025-05-14 writeup/paper.tex
minor fixes
almost done

@@
 \begin{abstract}
-A majority of large language models ingest word fragments called \textit{tokens} produced by a data compression algorithm known as byte pair encoding.
-This algorithm groups high-frequency letter pairings in natural language into individual units.
+A majority of large language models ingest word fragments produced by a data compression algorithm known as byte pair encoding.
+This algorithm groups high-frequency letter pairings in natural language into individual units called  \textit{tokens}.
 A natural question is whether the tokens produced by this pairing process deviate significantly from the source language's frequency distribution.
@@
 \section{Introduction}
+\label{section:intro}
 
 George Zipf in his hallmark \textit{The Psycho-Biology of Language} \cite{Zip35} introduced the remarkable trend that the word frequencies in many human languages exhibit the same power law distribution.
-Specifically, Zipf said that the frequency  that a word appears in some large corpus is proportional to the word's frequency rank.
+Specifically, Zipf said the number of times a word appears in a corpus is proportional to the word's frequency rank.
 Mathematically, Zipf's laws states
@@
 The accuracy of the trend in Equation \ref{eq:zipfmandelbrot} has been examined in $10^8$ English words in \cite{cancho2000}, over 50 languages in \cite{yu2018zipfslaw50languages}, and written-versus-spoken corpuses in \cite{lin2015scalinglawshumanspeech}.
-Each of these studies demonstrate that Zipf's law is generally exhibitted for common and somewhat-uncommon words, but rare words (words with high rank) appear less frequently than predicted.
-This deviation is shown to be statistically significant, and appears as two trendlines in the log-log plot.
-In English corpuses, this transition is found around the $10^5$\textit{-th} ranked word.\footnote{Some of these studies also find that extremely common words appear slighly more commonly than the Zipfian prediction.} The authors further explore the linguistic relavence and universality of these multiple regimes of words.
+Each of these studies demonstrate that Zipf's law is generally exhibitted for common and somewhat-uncommon words, but rare words (words with high rank) appear less frequently than predicted.\footnote{Some of these studies also find that extremely common words appear slighly more commonly than the Zipfian prediction.} This deviation is shown to be statistically significant, and appears as two trendlines in the log-log plot.
+In English corpuses, this transition is found around the $10^5$\textit{-th} ranked word.
+The authors further explore the linguistic relevance and universality of these multiple regimes of words.
 
@@
 However, unlike humans, LLMs digest text using a fixed vocabulary of word fragments called \textit{tokens}.
-The mapping between natural language and tokens is most commonly computed using the \textit{byte pair encoding algorithm} \cite{bpegage} over some large training corpus.
+The mapping between natural language and tokens is most commonly computed using the \textit{byte pair encoding (BPE) algorithm} \cite{bpegage} over some large training corpus.
 Byte pair encoding identifies the most frequent pair of letters, and creates a new token in the "token vocabulary" to replace that pair.
 Iterating this procedure until some fixed vocabulary size is reached creates the vocabulary of tokens.
-Finally, any sequence that does not map to one of the derived tokens is labelled with the \texttt{<unk>} token, standing for unknown.
+Finally, any sequence that does not map to one of the derived tokens is labelled with the ``unknown'' token.
 
@@
 
-To explore this question, we choose two modern tokenizers to analyze.
+To explore this question, we examine two modern tokenizers.
 We choose the tokenizer for the RoBERTa language model \cite{liu2019robertarobustlyoptimizedbert}, as the training data is publicly known.
 This allows us to conduct frequency analyses on corpuses that are on and off-distribution for the tokenizer.
-We also choose to compare our results with the tokenizer for the GPT-4 language model \cite{openai2024gpt4technicalreport} as a representative of an industry-grade tokenizer.
+We also choose to compare our results with the tokenizer for the GPT-4 language model \cite{openai2024gpt4technicalreport} as a representative of industry-scale byte pair encoding.
 
 We choose two corpuses based on RoBERTa's training data.
-We choose the 4.4GB bookscorpus dataset \cite{zhu2015aligningbooksmoviesstorylike}, as RoBERTa's tokenizer was trained (in-part) on this corpus.
-For our off-distribution corpus, we choose the 5.6GBtying MiniPile dataset \cite{kaddour2023minipilechallengedataefficientlanguage}, which was released after the RoBERTa model was released.
+We choose the 4.4GB bookscorpus dataset \cite{zhu2015aligningbooksmoviesstorylike}, which is within RoBERTa's training data, and the 5.6GB MiniPile dataset \cite{kaddour2023minipilechallengedataefficientlanguage}, which is not.
 Note that it is likely that the GPT-4 tokenizer was trained (in-part) on both of these corpuses.
@@
 We compute the word and token frequency for both tokenizers on each corpus, where we define a word as anything separated by a space.
-We then compute the data's Kolmogorov-Smirnov goodness-of-fit statistic for the Zipf distribution, and compare to that for an exponential and log-normal distribution.
-We do not remove any tokens from consideration unless otherwise stated.
+We then compute the data's Kolmogorov-Smirnov goodness-of-fit statistic for the fitted Zipf distribution, and compare to that for a fitted exponential and log-normal distribution.
+
+We do not remove any words or tokens from consideration unless otherwise stated, as to a language model all non-padding tokens are treated the same in training.
 
@@
 We find that token vocabularies are remarkably Zipfian, excluding extremely rare tokens.
-This holds true for both tokenizers and both corpuses studied, including out-of-distribution text.
-In the log-log plots in Figure \ref{fig:experiments}, we show sharp declines from the Zipfian trend only for the final few tokens in the RoBERTa tokenizer for both corpuses and the GPT-4 tokenizer for the MiniPile corpus.
-However, the GPT-4 tokenizer trend on the bookscorpus seems to deviate more dramatically, similar to the multiple regime studies.
+This holds true for both tokenizers and both corpuses studied, including the out-of-distribution text.
+In the log-log plots in Figure \ref{fig:experiments}, we find a clear linear trend over a large number of words and tokens.
+We only see deviations from the Zipfian trend for the final few tokens in the RoBERTa tokenizer for both corpuses and the GPT-4 tokenizer for the MiniPile corpus.
+However, the GPT-4 tokenizer trend on the bookscorpus seems to deviate more dramatically, similar to the multiple regime studies mentioned in Section \ref{section: intro}.
 
@@
 
-The lowest rank words and tokens for each tokenizer and each corpus, including control tokens and punctuation, are summarized in Table \ref{tbl:token ranks}.
+We also report the lowest rank words and tokens for each tokenizer and corpus to identify trends across models and corpuses.
+We report these words and tokens both including and excluding control tokens and punctuation, which are summarized in Table \ref{tbl:token ranks} and Table \ref{tbl:token ranks no control} respectively.
 
@@
         \hline
-        \textbf{Rank} & \multicolumn{2}{c|}{\textbf{Words}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4}}  \\
+        \textbf{Rank} & \multicolumn{2}{c|}{\textbf{Word}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4}}  \\
         \hline
@@
     \end{tabular}
-    \caption{Summary of most common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn MiniPile, \textit{not} including control tokens and punctuation}
+    \caption{Summary of most common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn MiniPile, \textit{not} including control tokens and punctuation.
+A majority of these words are considered \textit{function words} in linguistics literature.}
     \label{tbl:token ranks no control}
@@
 
-The Kolmogorov-Smirnov test statistics for each considered distribution are summarized in Table \ref{tbl:kstest}.
+Finally, we report the Kolmogorov-Smirnov test statistics for fitted Zipf-Mandelbrot, Exponential, and log-normal distributions in Table \ref{tbl:kstest}.
 
@@
         \hline
-        Exponential & 0.874 & 0.808 & 0.580 & 0.482 & 0.718 & 0.563 \\
+        Exponential & 0.87 & 0.81 & 0.58 & 0.48 & 0.72 & 0.56 \\
         \hline
-        log-normal & 0.379 & 0.451 & 0.032 & 0.050 & 0.397 & 0.283 \\
+        log-normal & 0.38 & 0.45 & 0.03 & 0.05 & 0.40 & 0.28 \\
         \hline
@@
 
-\section{Conclusions}
+\section{Analysis}
 
-Despite
+Despite the BPE algorithm building a vocabulary by directly referencing the frequency of subwords, these vocabularies are still Zipfian, and follow the word frequency distribution up to the rarest few tokens.
+Moreover, the tokenizers this algorithm produces creates Zipfian distributed tokens outside of the corpus it is trained on.
+When considering all possible tokens, the most common tokens tend to be punctuation, newline characters, and control tokens.
+Excluding punctuation and special tokens, we recover the known most common English words.
 
-\section{Future Work}
+\section{Conclusions \& Future Work}
+
+Noticeable theoretical work has been done in service of explaining Zipf's law for language \cite{li1992}
+
+\textit{On the statistical laws of linguistic distributions}\cite{Belevitch1959}
+
+Are a
 

### 2025-05-14 writeup/paper.tex
whoops

@@
 
-Despite the BPE algorithm building a vocabulary by directly referencing the frequency of subwords, these vocabularies are still Zipfian, and follow the word frequency distribution up to the rarest few tokens.
+The BPE algorithm builds a finite vocabulary by directly referencing the frequency of subwords and maps everything not captured to a "catch-all" unknown token.
+Despite this, these generated vocabularies are still Zipfian up to the rarest few tokens.
 Moreover, the tokenizers this algorithm produces creates Zipfian distributed tokens outside of the corpus it is trained on.
@@
 
-Noticeable theoretical work has been done in service of explaining Zipf's law for language \cite{li1992}
-
-\textit{On the statistical laws of linguistic distributions}\cite{Belevitch1959}
-
-Are a
-
-TODO
-
-How do token vocabularies differ from english vocabularies? Fixed vocabulary size, catchall token, functional tokens
+Noticeable theoretical work has been done in service of explaining Zipf's law for language \cite{li1992}.
+Most significant to the discussion was Belecitch's \textit{On the statistical laws of linguistic distributions}\cite{Belevitch1959}, in which the author shows the first order approximation of the rank ordering of many statistical distributions are all Zipfian.
+Can this approximation be shown to be invariant under the byte pair encoding transformation?
 

### 2025-05-14 writeup/paper.tex
wow its late

@@
     \caption{Summary of most common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn MiniPile, \textit{not} including control tokens and punctuation.
-A majority of these words are considered \textit{function words} in linguistics literature.}
+All of these words are considered \textit{function words} in linguistics literature.}
     \label{tbl:token ranks no control}
@@
 
-\begin{table}
-    \centering
-    \begin{tabular}{|c|c|c|c|c|c|c|}
-        \hline
-        \textbf{Distribution} & \multicolumn{2}{c|}{\textbf{Words}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4}}  \\
-        \hline
-        Zipf &  &  &  &  &  &  \\
-        \hline
-        Exponential & 0.87 & 0.81 & 0.58 & 0.48 & 0.72 & 0.56 \\
-        \hline
-        log-normal & 0.38 & 0.45 & 0.03 & 0.05 & 0.40 & 0.28 \\
-        \hline
-    \end{tabular}
-    \caption{Kolmogorov-Smirnov test statistics for distribution fit, where the left subcolumn is bookscorpus and the right subcolumn MiniPile}
-    \label{tbl:kstest}
-\end{table}
+% \begin{table}
+%     \centering
+%     \begin{tabular}{|c|c|c|c|c|c|c|}
+%         \hline
+%         \textbf{Distribution} & \multicolumn{2}{c|}{\textbf{Words}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4}}  \\
+%         \hline
+%         Zipf &  &  &  &  &  &  \\
+%         \hline
+%         Exponential & 0.87 & 0.81 & 0.58 & 0.48 & 0.72 & 0.56 \\
+%         \hline
+%         log-normal & 0.38 & 0.45 & 0.03 & 0.05 & 0.40 & 0.28 \\
+%         \hline
+%     \end{tabular}
+%     \caption{Kolmogorov-Smirnov test statistics for distribution fit, where the left subcolumn is bookscorpus and the right subcolumn MiniPile}
+%     \label{tbl:kstest}
+% \end{table}
 
@@
 
-The BPE algorithm builds a finite vocabulary by directly referencing the frequency of subwords and maps everything not captured to a "catch-all" unknown token.
+The BPE algorithm builds a finite vocabulary by directly referencing the frequency of subwords and maps everything not captured to a ``catch-all" unknown token.
 Despite this, these generated vocabularies are still Zipfian up to the rarest few tokens.
@@
 Noticeable theoretical work has been done in service of explaining Zipf's law for language \cite{li1992}.
-Most significant to the discussion was Belecitch's \textit{On the statistical laws of linguistic distributions}\cite{Belevitch1959}, in which the author shows the first order approximation of the rank ordering of many statistical distributions are all Zipfian.
+Most significant was Belecitch's \textit{On the statistical laws of linguistic distributions}\cite{Belevitch1959}, in which the author shows the first order approximation of the rank ordering of many statistical distributions are all Zipfian.
 Can this approximation be shown to be invariant under the byte pair encoding transformation?

### 2025-05-14 writeup/paper.tex
added belevich paper

@@
 Noticeable theoretical work has been done in service of explaining Zipf's law for language \cite{li1992}.
-Most significant was Belecitch's \textit{On the statistical laws of linguistic distributions}\cite{Belevitch1959}, in which the author shows the first order approximation of the rank ordering of many statistical distributions are all Zipfian.
-Can this approximation be shown to be invariant under the byte pair encoding transformation?
+Most significant was Belevitch's \textit{On the statistical laws of linguistic distributions}\cite{Belevitch1959}, in which the author shows the first order approximation of the rank ordering of many statistical distributions are all Zipfian.
+This indicates that Zipf's law may be due to the rank ordering of words more than the underlying formation of language, which is what Zipf was originally interested in.
+Can Belevitch's proof be shown to be invariant under the byte pair encoding transformation?
 

### 2025-05-17 writeup/paper.tex
edits
updates

@@
 
-\author{Jack Hanke \\
-  Northwestern University \\
-  \And
-  Daniel Plotkin \\
-  Northwestern University \\
-  \And
-  Nicole Birova \\
-  Northwestern University}
+\author{Jack Hanke \qquad Daniel Plotkin \\ 
+        {\bf Nicole Birova} \qquad {\bf David Demeter} \\
+        Northwestern University }
+
+% \author{Jack Hanke \\
+%   Northwestern University \\
+%   \And
+%   Daniel Plotkin \\
+%   Northwestern University \\
+%   \And
+%   Nicole Birova \\
+%   Northwestern University \\
+%   \And
+%   David Demeter \\
+%   Northwestern University}
 
@@
 
-George Zipf in his hallmark \textit{The Psycho-Biology of Language} \cite{Zip35} introduced the remarkable trend that the word frequencies in many human languages exhibit the same power law distribution.
-Specifically, Zipf said the number of times a word appears in a corpus is proportional to the word's frequency rank.
+George Zipf in his hallmark \textit{The Psycho-Biology of Language} \cite{Zip35} introduced the trend that the word frequencies in many human languages exhibit the same power law distribution.
+Specifically, Zipf said the number of times a word appears in a corpus is proportional to the word's rank.
 Mathematically, Zipf's laws states
@@
 The accuracy of the trend in Equation \ref{eq:zipfmandelbrot} has been examined in $10^8$ English words in \cite{cancho2000}, over 50 languages in \cite{yu2018zipfslaw50languages}, and written-versus-spoken corpuses in \cite{lin2015scalinglawshumanspeech}.
-Each of these studies demonstrate that Zipf's law is generally exhibitted for common and somewhat-uncommon words, but rare words (words with high rank) appear less frequently than predicted.\footnote{Some of these studies also find that extremely common words appear slighly more commonly than the Zipfian prediction.} This deviation is shown to be statistically significant, and appears as two trendlines in the log-log plot.
-In English corpuses, this transition is found around the $10^5$\textit{-th} ranked word.
+Each of these studies demonstrate that Zipf's law is generally exhibitted for common and somewhat-uncommon words, but rare words (words with high rank) appear less frequently than predicted.\footnote{Some of these studies also find that extremely common words appear slighly more commonly than the Zipfian prediction.} This deviation is shown to be statistically significant, and appears as two linear trends in log-log rank frequency plots.
+This transition is found around the $10^4$\textit{-th} ranked word \cite{cancho2000} \cite{yu2018zipfslaw50languages}.
 The authors further explore the linguistic relevance and universality of these multiple regimes of words.
@@
 However, unlike humans, LLMs digest text using a fixed vocabulary of word fragments called \textit{tokens}.
-The mapping between natural language and tokens is most commonly computed using the \textit{byte pair encoding (BPE) algorithm} \cite{bpegage} over some large training corpus.
-Byte pair encoding identifies the most frequent pair of letters, and creates a new token in the "token vocabulary" to replace that pair.
-Iterating this procedure until some fixed vocabulary size is reached creates the vocabulary of tokens.
-Finally, any sequence that does not map to one of the derived tokens is labelled with the ``unknown'' token.
+The mapping between natural language and tokens is most commonly computed using the \textit{byte pair encoding (BPE) algorithm} introduced by Gage \cite{bpegage} and popluarized in natural laguage processing by \cite{sennrich2016neuralmachinetranslationrare}.
+Modern tokenizers are \textit{byte-level} \cite{radford2019language}, where the indivisible units of natural language are bytes in the Unicode representation of a specific character.
 
@@
 
-To explore this question, we examine two modern tokenizers.
-We choose the tokenizer for the RoBERTa language model \cite{liu2019robertarobustlyoptimizedbert}, as the training data is publicly known.
-This allows us to conduct frequency analyses on corpuses that are on and off-distribution for the tokenizer.
-We also choose to compare our results with the tokenizer for the GPT-4 language model \cite{openai2024gpt4technicalreport} as a representative of industry-scale byte pair encoding.
+To explore this question, we examine two modern byte-level BPE tokenizers.
+We choose the tokenizer for the RoBERTa language model \cite{liu2019robertarobustlyoptimizedbert}, which has a vocabulary size of 50,265.
+As the training data is publicly known, this allows us to conduct frequency analyses on corpuses that are on and off-distribution for the tokenizer.
+We also choose to compare our results with the 200,019 vocab GPT-4o tokenizer \cite{openai2024gpt4technicalreport}, as a representative of industry-scale byte pair encoding.
 
-We choose two corpuses based on RoBERTa's training data.
-We choose the 4.4GB bookscorpus dataset \cite{zhu2015aligningbooksmoviesstorylike}, which is within RoBERTa's training data, and the 5.6GB MiniPile dataset \cite{kaddour2023minipilechallengedataefficientlanguage}, which is not.
-Note that it is likely that the GPT-4 tokenizer was trained (in-part) on both of these corpuses.
+We examine the 4.4GB bookscorpus dataset \cite{zhu2015aligningbooksmoviesstorylike}, which is within RoBERTa's training data, and the 5.6GB MiniPile dataset \cite{kaddour2023minipilechallengedataefficientlanguage}, which is not.
+Note that it is likely that the GPT-4o tokenizer was trained (in-part) on both of these corpuses.
 
-We compute the word and token frequency for both tokenizers on each corpus, where we define a word as anything separated by a space.
-We then compute the data's Kolmogorov-Smirnov goodness-of-fit statistic for the fitted Zipf distribution, and compare to that for a fitted exponential and log-normal distribution.
+We compute the word and token frequency for both tokenizers on each corpus, where we define a word as anything separated by the space character (U+0020).
+We then compute the data's Kolmogorov-Smirnov goodness-of-fit statistic for the fitted Zipf distribution, and compare to that for fitted exponential and log-normal distributions.
 
@@
 
-We find that token vocabularies are remarkably Zipfian, excluding extremely rare tokens.
-This holds true for both tokenizers and both corpuses studied, including the out-of-distribution text.
-In the log-log plots in Figure \ref{fig:experiments}, we find a clear linear trend over a large number of words and tokens.
-We only see deviations from the Zipfian trend for the final few tokens in the RoBERTa tokenizer for both corpuses and the GPT-4 tokenizer for the MiniPile corpus.
-However, the GPT-4 tokenizer trend on the bookscorpus seems to deviate more dramatically, similar to the multiple regime studies mentioned in Section \ref{section: intro}.
+We find that token vocabularies are remarkably Zipfian, excluding extremely rare tokens. This holds true for both tokenizers and both corpuses studied, including the out-of-distribution text.
+
+For bookscorpus, we find 1.3M distinct words, 31,729 distinct RoBERTa tokens, and 12,859 distinct GPT-4o tokens.
+For MiniPile, we find 32M distinct words, 50,165 distinct RoBERTa tokens, and 178,416 distinct GPT-4o tokens.
+
+We plot the log-log rank frequencies for each corpus in Figure \ref{fig:experiments}.
+We find a clear linear trend over a large number of words and tokens.
+We only see deviations from the Zipfian trend for the final few tokens in the RoBERTa tokenizer for both corpuses and the GPT-4o tokenizer for the MiniPile corpus.
+However, the GPT-4o tokenizer trend on the bookscorpus seems to deviate more dramatically, similar to the multiple regime studies mentioned in Section \ref{section:intro}.
 
@@
 
-\begin{table}
+\begin{table*}
     \centering
@@
         \hline
-        \textbf{Rank} & \multicolumn{2}{c|}{\textbf{Word}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4}}  \\
+        \textbf{Rank} & \multicolumn{2}{c|}{\textbf{Word}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4o}}  \\
         \hline
@@
         \hline
+        -1 & restain & RootDir, & seq & ␣Archdemon & wares & aryny \\
+        \hline
+        -1 & liarliar & homocystinemia & okemon & ␣petertodd & slan & verlening \\
+        \hline
+        -1 & shop-that & halfday & ython & ␣councill & wier & ␣myx \\
+        \hline
     \end{tabular}
-    \caption{Summary of most common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn MiniPile, including control tokens and punctuation}
+    \caption{Summary of most and least common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn is MiniPile, including control tokens and punctuation.
+A rank of -1 indicates a word or token that appears only once in the given corpus.
+The space character (Unicode U+0020) is denoted by the `␣' character.
+}
     \label{tbl:token ranks}
-\end{table}
+\end{table*}
 
@@
     \end{tabular}
-    \caption{Summary of most common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn MiniPile, \textit{not} including control tokens and punctuation.
+    \caption{Summary of most common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn is MiniPile, \textit{not} including control tokens and punctuation.
 All of these words are considered \textit{function words} in linguistics literature.}
@@
 
-% \begin{table}
-%     \centering
-%     \begin{tabular}{|c|c|c|c|c|c|c|}
-%         \hline
-%         \textbf{Distribution} & \multicolumn{2}{c|}{\textbf{Words}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4}}  \\
-%         \hline
-%         Zipf &  &  &  &  &  &  \\
-%         \hline
-%         Exponential & 0.87 & 0.81 & 0.58 & 0.48 & 0.72 & 0.56 \\
-%         \hline
-%         log-normal & 0.38 & 0.45 & 0.03 & 0.05 & 0.40 & 0.28 \\
-%         \hline
-%     \end{tabular}
-%     \caption{Kolmogorov-Smirnov test statistics for distribution fit, where the left subcolumn is bookscorpus and the right subcolumn MiniPile}
-%     \label{tbl:kstest}
-% \end{table}
+\begin{table*}
+    \centering
+    \begin{tabular}{|c|c|c|c|c|c|c|}
+        \hline
+        \textbf{Distribution} & \multicolumn{2}{c|}{\textbf{Words}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4o}}  \\
+        \hline
+        Zipf &  &  &  &  &  &  \\
+        \hline
+        Exponential & 0.87 & 0.81 & 0.58 & 0.48 & 0.72 & 0.56 \\
+        \hline
+        log-normal & 0.38 & 0.45 & 0.03 & 0.05 & 0.40 & 0.28 \\
+        \hline
+    \end{tabular}
+    \caption{Kolmogorov-Smirnov test statistics for distribution fit, where the left subcolumn is bookscorpus and the right subcolumn MiniPile}
+    \label{tbl:kstest}
+\end{table*}
 

### 2025-05-19 utils/anim.py
theory stuff

@@ -1,0 +1,75 @@
+import os
+import matplotlib.pyplot as plt
+import matplotlib.animation as animation
+import numpy as np
+
+
+def anim(
+        xs_data: list[np.array],
+        loglog_data: list[np.array], 
+        zipfs_data: list[np.array], 
+        error_data: list[np.array],
+        save_path: str = './visualizations/movie.gif', 
+    ):
+    
+    fig, axs = plt.subplots(1, 2)
+
+    # update function for animation
+    def update(frame):
+        # clear figure
+        axs[0].clear()
+        axs[1].clear()
+        # axs[0].tick_params(labelsize=10,bottom=False,left=False,labelbottom=False,labelleft=False)
+        axs[0].set_title(f'Rank Frequency Plot')
+        axs[0].set_xlabel(f'Rank')
+        axs[0].set_ylabel(f'Frequency')
+        # axs[1].tick_params(labelsize=10,bottom=False,left=False,labelbottom=False,labelleft=False)
+        axs[1].set_title(f'Zipf Error')
+        axs[1].set_xlabel(f'BPE Step')
+        axs[1].set_ylabel(f'Log MSE from Zipfian Prediction')
+
+        xs = xs_data[frame]
+        freqs = loglog_data[frame]
+        zipf = zipfs_data[frame]
+
+        # plot rank frequency
+        axs[0].plot(
+            xs, 
+            freqs,
+            color='green',
+        )
+        axs[0].plot(
+            xs, 
+            zipf,
+            linestyle='dashed',
+            color='gray',
+        )
+        axs[0].set_yscale(f'log')
+        axs[0].set_xscale(f'log')
+
+        # plot errors
+        axs[1].plot(
+            [i for i in range(frame)], 
+            error_data[:frame],
+            color='red',
+        )
+        axs[1].set_yscale(f'log')
+        # axs[1].axis('off')
+        axs[1].tick_params(labelleft=False)
+
+        # Set the title based on the frame number
+        fig.suptitle(f'BPE Step {frame}', fontsize=10)
+
+        return [fig]
+
+    # Create the animation
+    anim = animation.FuncAnimation(
+        fig, 
+        update, 
+        frames=len(loglog_data), 
+        interval=1,
+        blit=True
+    )
+
+    anim.save(save_path)
+    print(f'Saved animation at: {save_path}')    

### 2025-05-19 utils/bpe.py

@@ -1,0 +1,51 @@
+import numpy as np
+from copy import deepcopy
+# adapted from: https://github.com/karpathy/minbpe/tree/master
+
+def merge(ids, pair, idx):
+    """
+    In the list of integers (ids), replace all consecutive occurrences
+    of pair with the new integer token idx
+    Example: ids=[1, 2, 3, 1, 2], pair=(1, 2), idx=4 -> [4, 3, 4]
+    """
+    newids = []
+    i = 0
+    while i < len(ids):
+        # if not at the very last position AND the pair matches, replace it
+        if ids[i] == pair[0] and i < len(ids) - 1 and ids[i+1] == pair[1]:
+            newids.append(idx)
+            i += 2
+        else:
+            newids.append(ids[i])
+            i += 1
+    return newids
+
+def get_stats(ids, counts=None):
+    """
+    Given a list of integers, return a dictionary of counts of consecutive pairs
+    Example: [1, 2, 3, 1, 2] -> {(1, 2): 2, (2, 3): 1, (3, 1): 1}
+    Optionally allows to update an existing dictionary of counts
+    """
+    counts = {} if counts is None else counts
+    for pair in zip(ids, ids[1:]): # iterate consecutive elements
+        counts[pair] = counts.get(pair, 0) + 1
+    return counts
+
+# single step of BPE
+def one_step_of_bpe(
+        ids: np.array,
+        merges: dict,
+        alphabet_size: int 
+        ):
+    # count up the number of times every consecutive pair appears
+    stats = get_stats(ids)
+    # find the pair with the highest count
+    pair = max(stats, key=stats.get)
+    # mint a new token: assign it the next available id
+    idx = alphabet_size + 1
+    # replace all occurrences of pair in ids with idx
+    ids = merge(ids, pair, idx)
+    # save the merge
+    merges[pair] = idx
+
+    return deepcopy(ids), deepcopy(merges), idx

### 2025-05-19 utils/freq.py

@@ -1,0 +1,44 @@
+import numpy as np
+from tqdm import tqdm
+
+# 
+def increment_dict(dictionary: dict, key: str):
+    try:
+        dictionary[key] += 1
+    except KeyError:
+        dictionary[key] = 1
+
+# 
+def sort_dictionary(dictionary: dict):
+    return {k: v for k, v in sorted(dictionary.items(), key=lambda item: item[1], reverse=True)}
+
+# 
+# increment frequency for a given item (sentence) and a given tokenizer
+def increment_frequency(dictionary:dict, item: str):
+    for count, word in enumerate(item.split()):
+        increment_dict(dictionary=dictionary, key=word)
+    return count+1
+
+# 
+def normalize_dictionary(dictionary: dict, denominator: int):
+    return_dictionary = {}
+    for key, value in dictionary.items():
+        return_dictionary[key] = value/denominator
+    
+    return return_dictionary
+
+# takes array and 
+def freq(arr: np.array):
+    raw_freq_dict = {}
+    count = 0
+
+    # prog_bar = tqdm(arr)
+    for raw_item in arr:
+        item = str(raw_item)
+        count += increment_frequency(dictionary=raw_freq_dict, item=item)
+
+    # 
+    sorted_freq_dict = sort_dictionary(dictionary=raw_freq_dict)
+    frequencies = normalize_dictionary(dictionary=sorted_freq_dict, denominator=count)
+
+    return frequencies

### 2025-05-30 writeup/paper.tex
proper ks calcs
theory experiment

@@
 
-\title{Modern Byte Pair Tokenizers are Zipfian}
+\title{Empirical Evidence that Modern Byte Pair Tokenizers are Zipfian}
 
@@
 \begin{abstract}
-A majority of large language models ingest word fragments produced by a data compression algorithm known as byte pair encoding.
-This algorithm groups high-frequency letter pairings in natural language into individual units called  \textit{tokens}.
-A natural question is whether the tokens produced by this pairing process deviate significantly from the source language's frequency distribution.
-Zipf famously showed that many natural language's frequency distribution follows a power law, commonly known as Zipf's Law.
+A majority of large language models ingest word fragments produced by an algorithm known as byte pair encoding.
+This algorithm groups high-frequency element pairings in natural language into individual units called \textit{tokens}.
+A natural question is whether the frequency distribution of the tokens produced by this pairing process deviate significantly from the source language's.
+Zipf showed that many natural language's frequency distribution follows a power law, commonly known as Zipf's Law.
 We examine two modern tokenizer's adherence to Zipf's law at the token level.
-We demonstrate that these tokenizers are Zipfian on two corpuses, and speculate as to why this is.
+We provide empirical evidence that these tokenizers are Zipfian on two corpuses, and speculate as to why this is.
+Additionally, we give evidence that Zipfness is preserved over many individual steps of byte pair encoding.
 \end{abstract}
@@
 
-George Zipf in his hallmark \textit{The Psycho-Biology of Language} \cite{Zip35} introduced the trend that the word frequencies in many human languages exhibit the same power law distribution.
-Specifically, Zipf said the number of times a word appears in a corpus is proportional to the word's rank.
-Mathematically, Zipf's laws states
+George Zipf, in \textit{The Psycho-Biology of Language} \cite{Zip35} introduced the trend that the number of occurences of a word, the word's \textit{frequency}, in many human languages exhibit the same power law distribution.
+Specifically, Zipf said the work frequency in a corpus is proportional to the word's frequency rank, that being
 
@@
 
-A more descriptive model of word frequency that is more commonly referenced in linguistics literature is the Zipf-Mandelbrot distribution
+For example, Zipf's law predicts that the second most frequent word will be half as common as the most frequenct word.
+A more descriptive model of word frequency that is more commonly referenced in linguistics literature is the Zipf-Mandelbrot distribution, written
 
 \begin{equation}
-    \mbox{word frequency} \propto \frac{1}{(\mbox{word rank} + b)^a}.
+    \mbox{word frequency} \propto \frac{1}{(\mbox{word rank} + b)^a}, 
     \label{eq:zipfmandelbrot}
@@
 The accuracy of the trend in Equation \ref{eq:zipfmandelbrot} has been examined in $10^8$ English words in \cite{cancho2000}, over 50 languages in \cite{yu2018zipfslaw50languages}, and written-versus-spoken corpuses in \cite{lin2015scalinglawshumanspeech}.
-Each of these studies demonstrate that Zipf's law is generally exhibitted for common and somewhat-uncommon words, but rare words (words with high rank) appear less frequently than predicted.\footnote{Some of these studies also find that extremely common words appear slighly more commonly than the Zipfian prediction.} This deviation is shown to be statistically significant, and appears as two linear trends in log-log rank frequency plots.
+Each of these studies demonstrate that Zipf's law is generally exhibitted for common and somewhat-uncommon words, but rare words (words with high rank) appear less frequently than predicted.
+This deviation is shown to be statistically significant, and appears as two linear trends in log-log rank frequency plots.
 This transition is found around the $10^4$\textit{-th} ranked word \cite{cancho2000} \cite{yu2018zipfslaw50languages}.
-The authors further explore the linguistic relevance and universality of these multiple regimes of words.
+The authors further explore the linguistic relevance and universality of these multiple trends of words.
 
-Nearly a century after Zipf's discovery, large language models (LLMs) generate text comparable to human communication.
-However, unlike humans, LLMs digest text using a fixed vocabulary of word fragments called \textit{tokens}.
-The mapping between natural language and tokens is most commonly computed using the \textit{byte pair encoding (BPE) algorithm} introduced by Gage \cite{bpegage} and popluarized in natural laguage processing by \cite{sennrich2016neuralmachinetranslationrare}.
-Modern tokenizers are \textit{byte-level} \cite{radford2019language}, where the indivisible units of natural language are bytes in the Unicode representation of a specific character.
+Nearly a century after Zipf's discovery, large language models (LLMs) generate text comparable to human communication \cite{llmturingtest}.
+LLMs digest text using a fixed vocabulary of word fragments called \textit{tokens}.
+The mapping between natural language and tokens is most commonly computed using the \textit{byte pair encoding (BPE) algorithm} introduced by Gage \cite{bpegage} for file compression.
+An adjustment of BPE was popluarized for natural laguage processing by \cite{sennrich2016neuralmachinetranslationrare}, wherein instead of seeking to optimally compress a corpus of text, it instead seeks to map a corpus to a small fixed vocabulary.
+In this paper, we consider the natural language processing variant of the algorithm.
+For BPE, the \textit{elements} of a corpus are what portion of the text is paired in the first iteration of the algorithm.
+Elements can be words, phonemes, or letters.
+Modern tokenizers use the bytes of the Unicode representation as the elements, and so are called \textit{byte-level} tokenizers \cite{radford2019language}.
 
-When considering these ideas in tandem, a natural question arises: given a corpus that appears Zipfian, how does the tokenization process affect this trend?
+When considering these ideas in tandem, a natural question arises: given a corpus that appears Zipfian, how does the BPE tokenization process affect this trend?
 
@@
 As the training data is publicly known, this allows us to conduct frequency analyses on corpuses that are on and off-distribution for the tokenizer.
-We also choose to compare our results with the 200,019 vocab GPT-4o tokenizer \cite{openai2024gpt4technicalreport}, as a representative of industry-scale byte pair encoding.
+We also compare our results with the 200,019 vocab GPT-4o tokenizer \cite{openai2024gpt4technicalreport}, as a representative of industry-scale byte pair encoding.
 
 We examine the 4.4GB bookscorpus dataset \cite{zhu2015aligningbooksmoviesstorylike}, which is within RoBERTa's training data, and the 5.6GB MiniPile dataset \cite{kaddour2023minipilechallengedataefficientlanguage}, which is not.
-Note that it is likely that the GPT-4o tokenizer was trained (in-part) on both of these corpuses.
 
-We compute the word and token frequency for both tokenizers on each corpus, where we define a word as anything separated by the space character (U+0020).
-We then compute the data's Kolmogorov-Smirnov goodness-of-fit statistic for the fitted Zipf distribution, and compare to that for fitted exponential and log-normal distributions.
+We compute the byte, word, and token frequency for both tokenizers on each corpus, where we define a word as anything separated by the space character (U+0020).
+We then compute the data's Kolmogorov-Smirnov goodness-of-fit statistic for the fitted Zipf distribution, and compare the fit to fitted exponential and log-normal distributions.
+We do not remove any words or tokens from consideration unless otherwise stated.
 
-We do not remove any words or tokens from consideration unless otherwise stated, as to a language model all non-padding tokens are treated the same in training.
+Additionally, we create a synthetic dataset of $10^5$ samples from a Zipf distribution alphabet of size $100$ to examine how the distribution deviates from Zipf over successive steps of BPE.
 
@@
 
-We find that token vocabularies are remarkably Zipfian, excluding extremely rare tokens. This holds true for both tokenizers and both corpuses studied, including the out-of-distribution text.
+We find that token vocabularies are strongly Zipfian, excluding extremely rare tokens. This holds true for both tokenizers and both corpuses studied, including the out-of-distribution text.
 
@@
     \end{tabular}
-    \caption{Summary of most and least common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn is MiniPile, including control tokens and punctuation.
+    \caption{Summary of most and least common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn is MiniPile, \textit{including} control tokens and punctuation.
 A rank of -1 indicates a word or token that appears only once in the given corpus.
@@
     \end{tabular}
-    \caption{Summary of most common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn is MiniPile, \textit{not} including control tokens and punctuation.
+    \caption{Summary of most common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn is MiniPile, \textit{excluding} control tokens and punctuation.
 All of these words are considered \textit{function words} in linguistics literature.}
@@
 
-Finally, we report the Kolmogorov-Smirnov test statistics for fitted Zipf-Mandelbrot, Exponential, and log-normal distributions in Table \ref{tbl:kstest}.
+We report the Kolmogorov-Smirnov test statistics for fitted Zipf-Mandelbrot, Exponential, and log-normal distributions on $10^5$ samples from the rank-frequency data in Table \ref{tbl:kstest}.
 
@@
         \hline
-        Zipf &  &  &  &  &  &  \\
+        Zipf & 0.3324 & 03136 & 0.3307 & 0.3157 & 0.3785 & 0.3136 \\
         \hline
-        Exponential & 0.87 & 0.81 & 0.58 & 0.48 & 0.72 & 0.56 \\
+        Exponential & 0.5457 & 0.4405 & 0.5180 & 0.4728 & 0.5042 & 0.4287 \\
         \hline
-        log-normal & 0.38 & 0.45 & 0.03 & 0.05 & 0.40 & 0.28 \\
+        log-normal & 0.3708 & 0.3129 & 0.3675 & 0.3581 & 0.4230 & 0.2544 \\
         \hline
@@
 
+Figure \ref{fig:theoryexperiments} shows the mean squared error (MSE) between the Zipfian prediction and successive applications of a single step of BPE over the Zipfian synthetic dataset.
+
+\begin{figure}[t]
+  \includegraphics[width=\columnwidth]{../visualizations/zipfnessovertime.png}
+  \caption{Mean Squared Error from Zipf prediction of BPE applied to $10^5$ samples of a synthetic Zipf-distributed corpus of $100$ elements}
+  \label{fig:theoryexperiments}
+\end{figure}
+
 \section{Analysis}
 
-The BPE algorithm builds a finite vocabulary by directly referencing the frequency of subwords and maps everything not captured to a ``catch-all" unknown token.
-Despite this, these generated vocabularies are still Zipfian up to the rarest few tokens.
-Moreover, the tokenizers this algorithm produces creates Zipfian distributed tokens outside of the corpus it is trained on.
-When considering all possible tokens, the most common tokens tend to be punctuation, newline characters, and control tokens.
+We demonstrate a strong Zipfian trend in all tokenizers and corpuses studied, regardless of whether or not the tokenizer was trained on a given corpus.
+Deviations from the Zipf prediction only appear for the final few tokens.
+Over the token vocabulary, the most common tokens tend to be punctuation, newline characters, and control tokens.
 Excluding punctuation and special tokens, we recover the known most common English words.
 
+Additionally, synthetic data that is generated from a Zipfian distribution remain Zipfian after many successive steps of BPE.
+
 \section{Conclusions \& Future Work}
+
+The BPE algorithm, despite building vocabularies using the most frequent pairings of elements, generates vocabularies that are still Zipfian, up to the rarest few tokens.
 
@@
 This indicates that Zipf's law may be due to the rank ordering of words more than the underlying formation of language, which is what Zipf was originally interested in.
-Can Belevitch's proof be shown to be invariant under the byte pair encoding transformation?
+Can Belevitch's proof be shown to be invariant for some number of byte pair encoding transformations?
 

### 2025-05-30 utils/bpe.py

@@ -38,7 +38,7 @@ def one_step_of_bpe(
         alphabet_size: int 
         ):
     # count up the number of times every consecutive pair appears
-    stats = get_stats(ids)
+    stats = get_stats(ids=ids)
     # find the pair with the highest count
     pair = max(stats, key=stats.get)
     # mint a new token: assign it the next available id

### 2025-05-31 writeup/paper.tex
ok thats done yay

@@
 This algorithm groups high-frequency element pairings in natural language into individual units called \textit{tokens}.
-A natural question is whether the frequency distribution of the tokens produced by this pairing process deviate significantly from the source language's.
-Zipf showed that many natural language's frequency distribution follows a power law, commonly known as Zipf's Law.
+A natural question is whether the frequency distribution of the tokens produced by this pairing deviate significantly from the source language's.
+Zipf showed that many natural language's frequency distribution follow a power law, commonly known as Zipf's Law.
 We examine two modern tokenizer's adherence to Zipf's law at the token level.
@@
 
-George Zipf, in \textit{The Psycho-Biology of Language} \cite{Zip35} introduced the trend that the number of occurences of a word, the word's \textit{frequency}, in many human languages exhibit the same power law distribution.
-Specifically, Zipf said the work frequency in a corpus is proportional to the word's frequency rank, that being
+George Zipf, in \textit{The Psycho-Biology of Language} \cite{Zip35} introduced the trend that the number of occurences of a word, the word's \textit{frequency}, exhibits a power law.
+Specifically, Zipf said the word frequency in a corpus is proportional to the word's frequency rank, that being
 
@@
 This deviation is shown to be statistically significant, and appears as two linear trends in log-log rank frequency plots.
-This transition is found around the $10^4$\textit{-th} ranked word \cite{cancho2000} \cite{yu2018zipfslaw50languages}.
-The authors further explore the linguistic relevance and universality of these multiple trends of words.
+In English text, this transition is found around the $10^4$\textit{-th} ranked word \cite{cancho2000} \cite{yu2018zipfslaw50languages}.
+The authors further explore the linguistic relevance and universality of these multiple trends.
 
@@
 LLMs digest text using a fixed vocabulary of word fragments called \textit{tokens}.
-The mapping between natural language and tokens is most commonly computed using the \textit{byte pair encoding (BPE) algorithm} introduced by Gage \cite{bpegage} for file compression.
-An adjustment of BPE was popluarized for natural laguage processing by \cite{sennrich2016neuralmachinetranslationrare}, wherein instead of seeking to optimally compress a corpus of text, it instead seeks to map a corpus to a small fixed vocabulary.
+The mapping between natural language and tokens is most commonly computed using the \textit{byte pair encoding algorithm (BPE)}.
+BPE was introduced by Gage \cite{bpegage} for file compression, and later adapted for use in natural laguage processing by \cite{sennrich2016neuralmachinetranslationrare}.
+The NLP version of BPE, instead of seeking to optimally compress a corpus of text into a vocabulary of any size, instead maps a corpus to a small fixed vocabulary.
 In this paper, we consider the natural language processing variant of the algorithm.
-For BPE, the \textit{elements} of a corpus are what portion of the text is paired in the first iteration of the algorithm.
+
+BPE also requires specifying the \textit{elements} of a corpus, which are the portion of the text that is considered a indivisible chunk of the text.
 Elements can be words, phonemes, or letters.
-Modern tokenizers use the bytes of the Unicode representation as the elements, and so are called \textit{byte-level} tokenizers \cite{radford2019language}.
+Modern tokenizers use the bytes of the Unicode representation as the elements, and so are called \textit{byte-level} BPE tokenizers \cite{radford2019language}.
 
@@
 
-We compute the byte, word, and token frequency for both tokenizers on each corpus, where we define a word as anything separated by the space character (U+0020).
-We then compute the data's Kolmogorov-Smirnov goodness-of-fit statistic for the fitted Zipf distribution, and compare the fit to fitted exponential and log-normal distributions.
+We compute the word and token frequency for both tokenizers on each corpus, where we define a word as anything separated by the space character (U+0020).
+We then compute the data's $\chi^2$ goodness-of-fit statistic for the fitted Zipf distribution, and compare the fit to exponential and log-normal distributions.
 We do not remove any words or tokens from consideration unless otherwise stated.
 
-Additionally, we create a synthetic dataset of $10^5$ samples from a Zipf distribution alphabet of size $100$ to examine how the distribution deviates from Zipf over successive steps of BPE.
+Additionally, we create a synthetic dataset of $10^5$ samples from a zipf-distibuted alphabet of size $100$ to examine how the distribution deviates from Zipf over successive steps of BPE.
 
@@
 
-We find that token vocabularies are strongly Zipfian, excluding extremely rare tokens. This holds true for both tokenizers and both corpuses studied, including the out-of-distribution text.
-
 For bookscorpus, we find 1.3M distinct words, 31,729 distinct RoBERTa tokens, and 12,859 distinct GPT-4o tokens.
 For MiniPile, we find 32M distinct words, 50,165 distinct RoBERTa tokens, and 178,416 distinct GPT-4o tokens.
-
-We plot the log-log rank frequencies for each corpus in Figure \ref{fig:experiments}.
-We find a clear linear trend over a large number of words and tokens.
-We only see deviations from the Zipfian trend for the final few tokens in the RoBERTa tokenizer for both corpuses and the GPT-4o tokenizer for the MiniPile corpus.
-However, the GPT-4o tokenizer trend on the bookscorpus seems to deviate more dramatically, similar to the multiple regime studies mentioned in Section \ref{section:intro}.
+We plot the log-log rank frequency for both corpuses in Figure \ref{fig:experiments}.
 
@@
 
-We also report the lowest rank words and tokens for each tokenizer and corpus to identify trends across models and corpuses.
-We report these words and tokens both including and excluding control tokens and punctuation, which are summarized in Table \ref{tbl:token ranks} and Table \ref{tbl:token ranks no control} respectively.
+We also report the least and most common words and tokens for each tokenizer and corpus.
+We report these words and tokens both including and excluding control tokens and punctuation, which are summarized in Table \ref{tbl:token ranks} and Table \ref{tbl:token ranks no control}.
 
@@
 A rank of -1 indicates a word or token that appears only once in the given corpus.
-The space character (Unicode U+0020) is denoted by the `␣' character.
-}
+The space character (Unicode U+0020) is denoted by the `␣' character.}
     \label{tbl:token ranks}
@@
 
-We report the Kolmogorov-Smirnov test statistics for fitted Zipf-Mandelbrot, Exponential, and log-normal distributions on $10^5$ samples from the rank-frequency data in Table \ref{tbl:kstest}.
+We report the $\chi^2$ test statistics for fitted Zipf-Mandelbrot in Table \ref{tbl:kstest}, where we use Equation \ref{eq:zipfmandelbrot} with $a=1.1$ and $b=0$ for all Zipf tests.
 
@@
     \centering
-    \begin{tabular}{|c|c|c|c|c|c|c|}
+    \begin{tabular}{|c|c|c|c|}
         \hline
-        \textbf{Distribution} & \multicolumn{2}{c|}{\textbf{Words}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4o}}  \\
+        \textbf{Distribution} & \textbf{Words} & \textbf{RoBERTa} & \textbf{GPT-4o}  \\
         \hline
-        Zipf & 0.3324 & 03136 & 0.3307 & 0.3157 & 0.3785 & 0.3136 \\
+        bookscorpus & $2.5 \cdot 10^{9}$ & $1.6 \cdot 10^{9}$ & $1.4 \cdot 10^{8}$  \\
         \hline
-        Exponential & 0.5457 & 0.4405 & 0.5180 & 0.4728 & 0.5042 & 0.4287 \\
-        \hline
-        log-normal & 0.3708 & 0.3129 & 0.3675 & 0.3581 & 0.4230 & 0.2544 \\
+        MiniPile &  $1.4 \cdot 10^{9}$ & $4.1 \cdot 10^{9}$ & $1.8 \cdot 10^{9}$ \\
         \hline
     \end{tabular}
-    \caption{Kolmogorov-Smirnov test statistics for distribution fit, where the left subcolumn is bookscorpus and the right subcolumn MiniPile}
+    \caption{$\chi^2$ test statistics for distribution fit for each tokenizer and corpus, each of which have a p value of near $0$}
     \label{tbl:kstest}
@@
 
-Figure \ref{fig:theoryexperiments} shows the mean squared error (MSE) between the Zipfian prediction and successive applications of a single step of BPE over the Zipfian synthetic dataset.
+Finally, we compute the mean squared error (MSE) between the Zipfian prediction and successive applications of a single step of BPE over the Zipfian synthetic dataset in Figure \ref{fig:theoryexperiments}.
 
@@
 
-We demonstrate a strong Zipfian trend in all tokenizers and corpuses studied, regardless of whether or not the tokenizer was trained on a given corpus.
-Deviations from the Zipf prediction only appear for the final few tokens.
+In general, we find strong evidence that BPE tokenizers are Zipfian.
+Deviations from the Zipfian trend only appear for the final few RoBERTa tokens for both corpuses and the GPT-4o tokens for the MiniPile corpus.
+This includes in and out-of-distribution text.
+However, the GPT-4o tokenizer trend on the bookscorpus seems to deviate more dramatically, similar to the multiple regime studies mentioned in Section \ref{section:intro}.
+We speculate that this may be due to Minipile being a multilingual dataset, which may affect the the distribution for a multilingual tokenizer such as GPT-4o.
+The $\chi^2$ goodness-of-fit tests confirm a Zipf describes the data well, all having p-values $<<0.0001$.
+
 Over the token vocabulary, the most common tokens tend to be punctuation, newline characters, and control tokens.
@@
 
-Additionally, synthetic data that is generated from a Zipfian distribution remain Zipfian after many successive steps of BPE.
+Additionally, applying the BPE pairing process to the synthetic data shows almost no deviations from the original Zipf distribution for hundreds of iterations.
+However, after about $700$ iterations the MSE from Zipf does begin to increase, though the error is still relatively small.
 
@@
 
-The BPE algorithm, despite building vocabularies using the most frequent pairings of elements, generates vocabularies that are still Zipfian, up to the rarest few tokens.
+The byte pair encoding algorithm, despite building vocabularies using the most frequent pairings of elements, generates vocabularies that are Zipfian, up to the rarest few tokens.
+We demonstate this over multiple tokenizers and corpuses, including both in and out-of-distribution text.
+We also provide empirical evidence that Zipfness is preserved over many iterations of BPE.
 
-Noticeable theoretical work has been done in service of explaining Zipf's law for language \cite{li1992}.
+Noticeable theoretical work has been done to explain Zipf's law for language \cite{li1992}.
 Most significant was Belevitch's \textit{On the statistical laws of linguistic distributions}\cite{Belevitch1959}, in which the author shows the first order approximation of the rank ordering of many statistical distributions are all Zipfian.
-This indicates that Zipf's law may be due to the rank ordering of words more than the underlying formation of language, which is what Zipf was originally interested in.
+This indicates that Zipf's law may be due to the rank ordering of words more than the underlying formation of language.
 Can Belevitch's proof be shown to be invariant for some number of byte pair encoding transformations?

### 2025-06-08 writeup/paper.tex
Fixed grammer and spelling errors

@@
 This algorithm groups high-frequency element pairings in natural language into individual units called \textit{tokens}.
-A natural question is whether the frequency distribution of the tokens produced by this pairing deviate significantly from the source language's.
-Zipf showed that many natural language's frequency distribution follow a power law, commonly known as Zipf's Law.
-We examine two modern tokenizer's adherence to Zipf's law at the token level.
-We provide empirical evidence that these tokenizers are Zipfian on two corpuses, and speculate as to why this is.
+A natural question is whether the frequency distribution of the tokens produced by this pairing deviate significantly from the source language's distribution Zipf showed that many natural languages' frequency distribution follow a power law, commonly known as Zipf's Law.
+We examine two modern tokenizers' adherence to Zipf's law at the token level.
+We provide empirical evidence that these tokenizers are Zipfian on two corpora, and speculate as to why this is.
 Additionally, we give evidence that Zipfness is preserved over many individual steps of byte pair encoding.
@@
 
-George Zipf, in \textit{The Psycho-Biology of Language} \cite{Zip35} introduced the trend that the number of occurences of a word, the word's \textit{frequency}, exhibits a power law.
+George Zipf, in \textit{The Psycho-Biology of Language} \cite{Zip35} introduced the trend that the number of occurrences of a word, the word's \textit{frequency}, exhibits a power law.
 Specifically, Zipf said the word frequency in a corpus is proportional to the word's frequency rank, that being
@@
 
-For example, Zipf's law predicts that the second most frequent word will be half as common as the most frequenct word.
+For example, Zipf's law predicts that the second most frequent word will be half as common as the most frequent word.
 A more descriptive model of word frequency that is more commonly referenced in linguistics literature is the Zipf-Mandelbrot distribution, written
@@
 The accuracy of the trend in Equation \ref{eq:zipfmandelbrot} has been examined in $10^8$ English words in \cite{cancho2000}, over 50 languages in \cite{yu2018zipfslaw50languages}, and written-versus-spoken corpuses in \cite{lin2015scalinglawshumanspeech}.
-Each of these studies demonstrate that Zipf's law is generally exhibitted for common and somewhat-uncommon words, but rare words (words with high rank) appear less frequently than predicted.
+Each of these studies demonstrate that Zipf's law is generally exhibited for common and somewhat-uncommon words, but rare words (words with high rank) appear less frequently than predicted.
 This deviation is shown to be statistically significant, and appears as two linear trends in log-log rank frequency plots.
@@
 The mapping between natural language and tokens is most commonly computed using the \textit{byte pair encoding algorithm (BPE)}.
-BPE was introduced by Gage \cite{bpegage} for file compression, and later adapted for use in natural laguage processing by \cite{sennrich2016neuralmachinetranslationrare}.
+BPE was introduced by Gage \cite{bpegage} for file compression, and later adapted for use in natural language processing by \cite{sennrich2016neuralmachinetranslationrare}.
 The NLP version of BPE, instead of seeking to optimally compress a corpus of text into a vocabulary of any size, instead maps a corpus to a small fixed vocabulary.
@@
 
-BPE also requires specifying the \textit{elements} of a corpus, which are the portion of the text that is considered a indivisible chunk of the text.
+BPE also requires specifying the \textit{elements} of a corpus, which are the portion of the text that is considered an indivisible chunk of the text.
 Elements can be words, phonemes, or letters.
@@
 
-Additionally, we create a synthetic dataset of $10^5$ samples from a zipf-distibuted alphabet of size $100$ to examine how the distribution deviates from Zipf over successive steps of BPE.
+Additionally, we create a synthetic dataset of $10^5$ samples from a zipf-distributed alphabet of size $100$ to examine how the distribution deviates from Zipf over successive steps of BPE.
 
@@
 However, the GPT-4o tokenizer trend on the bookscorpus seems to deviate more dramatically, similar to the multiple regime studies mentioned in Section \ref{section:intro}.
-We speculate that this may be due to Minipile being a multilingual dataset, which may affect the the distribution for a multilingual tokenizer such as GPT-4o.
+We speculate that this may be due to Minipile being a multilingual dataset, which may affect the distribution for a multilingual tokenizer such as GPT-4o.
 The $\chi^2$ goodness-of-fit tests confirm a Zipf describes the data well, all having p-values $<<0.0001$.
@@
 The byte pair encoding algorithm, despite building vocabularies using the most frequent pairings of elements, generates vocabularies that are Zipfian, up to the rarest few tokens.
-We demonstate this over multiple tokenizers and corpuses, including both in and out-of-distribution text.
+We demonstrate this over multiple tokenizers and corpuses, including both in and out-of-distribution text.
 We also provide empirical evidence that Zipfness is preserved over many iterations of BPE.
